---
title: "NDV_supplement with `A vector vaccine reduces transmission of Newcastle disease virus in commercial broiler chickens with maternally derived antibodies' "
author: "T. Tatar-Kis, E.A.J. Fischer, C. Cazaban, E. Kovacs, Z. G. Homonnay, V. Palya and J.A.(Arjan) Stegeman"
date: "10-7-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Used libraries and functions

#libraries

```{r}
options("getSymbols.warning4.0"=FALSE)
#add libraries
library(openxlsx)
library(ggplot2)
library(lme4)
library(survival)
```

# User defined functions

```{r}

#colour-blind colour palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

#function to reform data for glm
source("NDV_datamodel.R")

```

## Load and transform data

```{r loaddata}
#set directory with data equal to the working directory
data.dir <- getwd()
#read the data of the vaccination group

#data original CT values
ndvdata <- read.xlsx("Appendix_individualsCtvaluesandevaluation.xlsx", 
                 startRow =2,
                 colNames = T)

#cevadata.original <- read.xlsx(paste0(data.dir, "Appendix-individuals Ct values and evaluation.xlsx"),
 #                       sheet = "RAW (4)")
ndvdata<-fill.type.na.rename(ndvdata)
#data used for transmission 
#cevadata <- read.xlsx(paste0(data.dir, "DataNDVtrial_manual_Egil 20200317_Timea20200319.xlsx"),
 #                       sheet = "Input_seedernotchanged")

#transform to usable data
data.one.SIR <- reform.SIR.data(data = ndvdata,
                                sampleday.vars= paste0(c(1:14),".dpch"),
                                exclude.seeder = T )
data.one.SIR$Vaccinated <- factor(data.one.SIR$Vaccinated, levels = c("Yes","No"))
data.one.SIR$Vaccinated<-relevel(data.one.SIR$Vaccinated, ref = "No")

```

## Excretion

For those that were defined positive

```{r include = F}
ct.data <- ndvdata
for(day in c(6:19)){
  
  ct.data[(!data.one.SIR[[2]][,day]==1)|is.na(data.one.SIR[[2]][,day]),day]<- NA
  
}

suppressWarnings(
 ct.data$AUC <- sapply(c(1:length(ct.data$Group)), function(j){sum(36-as.numeric(ct.data[j,6:19]), na.rm = T)}) )#do not bother about NA's

#selection only the ct values >0
ct.data<- ct.data[ct.data$AUC>0,]
 
ctauc <- aggregate(formula = AUC ~ Vaccinated+ci+Sample,
          data = ct.data[ct.data$AUC>0,],
          FUN = "mean", rm.na = T)
ctauc$sd<-aggregate(formula = AUC ~ Vaccinated+ci+Sample,
          data = ct.data[ct.data$AUC>0,],
          FUN = "sd")$AUC
ctauc$n<-aggregate(formula = AUC ~ Vaccinated+ci+Sample,
          data = ct.data[ct.data$AUC>0,],
          FUN = "length")$AUC
#combine all categories
ct.data$category <- paste0(ct.data$Vaccinated,".",ct.data$ci)

#overall test
kruskal.test(formula = AUC ~ category,data = ct.data[ct.data$Sample == "ON swab",])
kruskal.test(formula = AUC ~ category,data = ct.data[ct.data$Sample == "CLO swab",])

#cloacal swabs contact versus seeders
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "CLO swab",], paired = F)

#oro-nasal swabs contact versus seeders
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "ON swab",], paired = F)

 
#test seed - contact per treatment CLO in vaccinated
 wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$Vaccinated =="Yes",], paired = F)
 
 #test seed - contact per treatment ON in vaccinated
 wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "ON swab"&ct.data$Vaccinated =="Yes",], paired = F)

#test  seed=contact CLO in unvaccinated
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$Vaccinated =="No",], paired = F)
 
#test  seed=contact ON in unvaccinated
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "ON swab"&ct.data$Vaccinated =="No",], paired = F)


#test vaccinated versus unvaccinated stratified by seeder-contact
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$ci =="seeder",])
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$ci =="contact",])

 
#test vaccinated versus unvaccinated stratified by seeder-contact
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "ON swab"&ct.data$ci =="seeder",])
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "ON swab"&ct.data$ci =="contact",])


#test vaccinated vs unvaccinated overall
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "CLO swab",])
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "ON swab",])


```


## Infectious period

# Data selection

```{r Tinf, echo=TRUE}
#get data to calculate infectious period
out.indiv.one.SIR<- data.one.SIR[[2]];
#sum all positive intervals
out.indiv.one.SIR$infper <- apply(data.one.SIR[[2]][,c(5:18)]==1, 1, sum, na.rm = T);
#remove 0 as these are not infected thus should not be considered
out.indiv.one.SIR[out.indiv.one.SIR$infper == 0,]$infper<- NA #zero means never been infected
```

# Plot

```{r include = T}
#make factor
out.indiv.one.SIR$Vaccinated<-factor(out.indiv.one.SIR$Vaccinated, levels = c("Yes","No"))

hist.one.chal <- ggplot(data = out.indiv.one.SIR) +
  geom_histogram(aes(x = infper, y = ..density.. , fill = Vaccinated), colour = "black",
                 position = "dodge",binwidth = 1, na.rm = T)+
  scale_fill_manual(values=c("Yes" = cbPalette[2],"No" = cbPalette[3]))+
  scale_x_continuous(breaks = c(0:10))+
 # ggtitle("Distribution of infectious period")+
  ylab("Proportion of excreting chickens")+
  xlab("Infectious period (days)") +
  #theme(panel.background = NULL,  legend.position = "none")+
  theme_bw()+
  facet_grid( ~ Challenge,
             labeller = labeller(
       # Vaccinated = c("No" = "Unvaccinated", "Yes" = "Vaccinated"),
        Challenge = c("contact" = "Contact", "seeder" = "Seeder")
    )) 
ggsave("InfectiousPeriod.tiff",hist.one.chal, dpi =720)

hist.one.chal
```

# Estimate infectious period

The infectious period \(T_{inf}\): Parametric survival analyses assuming Normal distribution.

```{r include =T}
#add censoring as variable to the data set
out.indiv.one.SIR$cens <- 1-as.numeric(out.indiv.one.SIR$`14.dpch` == 1)
out.indiv.one.SIR$cens[is.na(out.indiv.one.SIR$cens)] <- 1
out.indiv.one.SIR[out.indiv.one.SIR$cens==0,]
#Survival analyses on contact birds
#perform exponential survival analysis
surv.fit.one.SIR.exponential <- survival::survreg(survival::Surv(infper,cens)~ relevel(Vaccinated, ref= "No"), dist = "exponential",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)
#perform gaussian survival analysis 
surv.fit.one.SIR.gaussian <- survival::survreg(survival::Surv(infper,cens)~ relevel(Vaccinated, ref= "No") , dist = "gaussian",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)
AIC(surv.fit.one.SIR.exponential)
AIC(surv.fit.one.SIR.gaussian)
```

Based on AIC select gaussian model.

```{r}
#check if dropping vaccination changes the model
drop1(surv.fit.one.SIR.gaussian,test = "Chisq")

#print confidence intervals
summary(surv.fit.one.SIR.gaussian)
Tinfci<-confint(surv.fit.one.SIR.gaussian)

```

drop1(surv.fit.one.SIR.gaussian,test = "Chisq")
summary(surv.fit.one.SIR.gaussian.sc)

# Calculate beta and R0

```{r}
#pre select data without rows for S =0 or I = 0 or S = NA or C = NA
out.nona.one.SIR <- data.one.SIR[[1]][(!is.na(data.one.SIR[[1]]$S)&!is.na(data.one.SIR[[1]]$C) &data.one.SIR[[1]]$I>0 &data.one.SIR[[1]]$S>0),]
out.nona.one.SIR$Vaccinated <- factor(out.nona.one.SIR$Vaccinated, levels = c("No","Yes"))
out.nona.one.SIR$Vaccinated <- relevel(out.nona.one.SIR$Vaccinated, ref = "No")
#estimate beta 
fit.SIR <- glm(cbind(C, S-C) ~ relevel(Vaccinated,ref = "No") ,
                offset = log(I/N),
                family = binomial(link = "cloglog"), data = out.nona.one.SIR, 
                na.action = na.omit)

#calculate R0
alpha = 0.05 #conf.level
#beta
mean.logbeta <- cumsum(summary(fit.SIR)$coefficients[,1])
var.logbeta <- summary(fit.SIR)$coefficients[,2]^2
confint.logbeta<-  confint(fit.SIR)+c(0,mean.logbeta[1])

#infectious period
mean.inf.per <- cumsum(summary(surv.fit.one.SIR.gaussian)$coefficients[1:2])
confint.inf.per <- confint(surv.fit.one.SIR.gaussian)[1:2,]+c(0, mean.inf.per[1])
#get the variance from the profile likelihood
var.inf.per <- ((confint.inf.per-mean.inf.per)[,2]/qnorm(0.05/2))^2
#calculate log of the mean and the variance of the log of the mean. 
mean.log.inf.per <- log(mean.inf.per)
var.log.inf.per <- log(1+var.inf.per/(mean.inf.per)^2)

#calculate R
mean.logR <- mean.logbeta + mean.log.inf.per
var.logR <- var.logbeta + var.log.inf.per
cll.logR <- mean.logR - qnorm(1-0.05/2) * sqrt(var.logR)
cul.logR <- mean.logR + qnorm(1-0.05/2) * sqrt(var.logR)

#put an a table
data.table <- data.frame(Tinf = round(mean.inf.per,2),
          Tinfll = round(confint.inf.per[,1],2),
          Tinful = round(confint.inf.per[,2],2),
          beta = round(exp(mean.logbeta),2),
          betall = round(exp(confint.logbeta)[,1],2),
          betaul =round(exp(confint.logbeta)[,2],2),
          R0 = round(exp(mean.logR),2),
          R0ll = round(exp (cll.logR),2),
          R0ul = round(exp(cul.logR),2))
row.names(data.table)<- c("Unvaccinated","Vaccinated")
out.table <- knitr::kable(x = data.table,
             digits =2,
             format = "html",
             #row.names = c("Unvaccinated","Vaccinated")),
             col.names =c("Estimate","2.5%","97.5%",
                          "Estimate","2.5%","97.5%",
                          "Estimate","2.5%","97.5%"))
kableExtra::add_header_above(out.table,c("Treatment" =1,
                 "Tinf" = 3,
                 "beta" = 3,
                 "R0" = 3))

```
Probability of a major outbreak is 0 for $R_0<= 1$ and $1-1/R_0 $ otherwise. 

```{r}
#probability major outbreak
(1-1/data.table$R0 ) #negative means 0
(1-1/data.table$R0ll ) #negative means 0
(1-1/data.table$R0ul ) #negative means 0

```
The final size can still be calculated as the absolute number of infections by one initiall infected $\dfrac{1}{1-R_0}$ or the fraction of the population $z$ eventually infected $-\ln(1-z) = z R_0$ :
```{r}
#final size of the outbreak
final.size<- function(R0){if(R0<1){return(1/(1-R0))}else{
 return(uniroot(function(z){-R0*z-log(1-z)},lower = .1, upper =1)$root)
  }
}

sapply(data.table$R0,final.size) #first is fraction, second is number
sapply(data.table$R0ll,final.size) #first is fraction, second is number
sapply(data.table$R0ul,final.size) #both fractions


```
