---
title: "NDV_supplement"
author: "Egil Fischer"
date: "2-7-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Used libraries and functions

#libraries

```{r}
options("getSymbols.warning4.0"=FALSE)
#add libraries
library(openxlsx)
library(ggplot2)
library(lme4)
library(survival)
```

# User defined functions

```{r}

#colour-blind colour palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

#function to reform data for glm
#uses file with states(positive / negative) already determined manually from Ct value
 reform.SIR.data <- function(data, 
                            sampleday.vars,
                            exclude.seeder = T
 ){
   reform.data(data,
               sampleday.vars = sampleday.vars,
               exclude.seeder = exclude.seeder,
               SIR.state = T)
 }

#count infections based on each type of sample separately 
reform.data<- function(data, 
                       model = c("SIS","SIR","SI")[1],#transmission model default SIS model
                       inf.rule =1, #rule with number of consecutive positive samples to be determined infectious
                       rec.rule =1, #rule with number of consecutive negative samples to be determined recovered
                       sampleday.vars,
                       group.by = NULL, 
                       positive.if = c("max","min")[1],#max will use at least 1 positive sample, min all samples positive.
                       cut.off = 0,
                       exclude.seeder = T,
                       SIR.state = F#False measn that SIR status needs to be determined
                       )
  {
  
  if(!SIR.state){
  #start with a recoding data to binary yes/no
  binary.data <- data; binary.data[, sampleday.vars] <- lapply(binary.data[, sampleday.vars], 
                                                               FUN = function(x){ifelse(x < cut.off,1,0)});
  #group positive samples based on positive.if statement
  aggregate.data <- aggregate.data.frame(binary.data[,sampleday.vars], 
                                         by = list(binary.data$Group, 
                                                   binary.data$bird.id, 
                                                   binary.data$Vaccinated,
                                                   binary.data$ci),
                                         FUN = positive.if)
  #recode based on transmission model
  for(ani in c(1:length(aggregate.data[,1])))
  {
   
    aggregate.data[ani,sampleday.vars]<-SIR.state(as.vector(aggregate.data[ani,sampleday.vars][1,]),
                                                  model = model,
                                                  inf.rule = inf.rule,
                                                  rec.rule = rec.rule )
    
  }
  
  #change the colnames
  colnames(aggregate.data)[1:4]<- c("Group","bird.id","Vaccinated","Challenge")
  }
   else{
    aggregate.data <- data #already as input presented.
   }
  #aggregate to number of infections prior to time step TODO is to put this in a separate function
  id<-NULL; inf <- NULL; rec<- NULL;n <- NULL; cases <- NULL;susceptibles <- NULL;
  for(day in c(1:length(sampleday.vars)))
  {
    #determine the number of infectious at this day
    inf <- rbind(inf, data.frame(dpch = sampleday.vars[day],
                                 ndpch = day,
                                 aggregate.data.frame(aggregate.data[,sampleday.vars[day]], 
                                                      by = list(aggregate.data$Group, aggregate.data$Vaccinated), 
                                                      FUN = function(x){sum(x==1,na.rm = TRUE)})
    ))
    #determine the number of recoverd at this day
    rec <- rbind(rec, data.frame(dpch = sampleday.vars[day],
                                 ndpch = day,
                                 aggregate.data.frame(aggregate.data[,sampleday.vars[day]], 
                                                      by = list(aggregate.data$Group, aggregate.data$Vaccinated), 
                                                      FUN = function(x){sum(x==2, na.rm = TRUE)})
    ))
    #determine the number of birds at this day
    n <- rbind(n, data.frame(dpch = sampleday.vars[day],
                             ndpch = day,
                             aggregate.data.frame(aggregate.data[,sampleday.vars[day]], 
                                                  by = list(aggregate.data$Group, aggregate.data$Vaccinated), 
                                                  FUN = function(x) sum( !is.na(x) )) 
                             
    ))
    
    #determine the number of susceptibles
    if(exclude.seeder){
      susceptibles <- rbind(susceptibles, data.frame(dpch = sampleday.vars[day],
                               ndpch = day,
                               aggregate.data.frame(aggregate.data[aggregate.data$Challenge == "contact",sampleday.vars[day]], 
                                                    by = list(aggregate.data$Group[aggregate.data$Challenge == "contact"], 
                                                              aggregate.data$Vaccinated[aggregate.data$Challenge == "contact"]), 
                                                    FUN = function(x) {sum(x==0, na.rm = TRUE)}))) 
      
    }else
    {
      susceptibles <- rbind(susceptibles, data.frame(dpch = sampleday.vars[day],
                                                     ndpch = day,
                                                     aggregate.data.frame(aggregate.data[,sampleday.vars[day]], 
                                                                          by = list(aggregate.data$Group, aggregate.data$Vaccinated), 
                                                                          FUN = function(x) {sum(x==0, na.rm = TRUE)})))
    }
    
    #determine the number of new cases at this day
    if(day < length(sampleday.vars)){
      if(exclude.seeder){
      cases <- rbind(cases, data.frame(dpch = sampleday.vars[day],
                                       ndpch = day,
                                       cases =  aggregate.data.frame((
                                         aggregate.data[aggregate.data$Challenge == "contact",sampleday.vars[day+1]]-aggregate.data[aggregate.data$Challenge == "contact",sampleday.vars[day]])>0, 
                                         by = list(aggregate.data$Group[aggregate.data$Challenge == "contact"], aggregate.data$Vaccinated[aggregate.data$Challenge == "contact"]), 
                                         FUN = sum, na.rm = TRUE)   )      )
      }
      else
        {
        cases <- rbind(cases, data.frame(dpch = sampleday.vars[day],
                                         ndpch = day,
                                         cases =  aggregate.data.frame((
                                           aggregate.data[,sampleday.vars[day+1]]-aggregate.data[,sampleday.vars[day]])>0, 
                                           by = list(aggregate.data$Group, aggregate.data$Vaccinated), 
                                           FUN = sum, na.rm = TRUE)))
      }
    }else {
     
      #determine cases and potential exclude the seeders
      if(exclude.seeder){
        cases <- rbind(cases, 
                       data.frame(dpch = sampleday.vars[day],
                                  ndpch = day,
                                  cases = aggregate.data.frame(aggregate.data[aggregate.data$Challenge == "contact",sampleday.vars[day]], 
                                                               by = list(aggregate.data$Group[aggregate.data$Challenge == "contact"],
                                                                         aggregate.data$Vaccinated[aggregate.data$Challenge == "contact"]), 
                                                               FUN = function(x){0}))
        )}
      else{
      cases <- rbind(cases, 
                     data.frame(dpch = sampleday.vars[day],
                                ndpch = day,
                                cases = aggregate.data.frame(aggregate.data[,sampleday.vars[day]], 
                                                             by = list(aggregate.data$Group, aggregate.data$Vaccinated), 
                                                             FUN = function(x){0}))
      )}
    }
  }
    
  #combine infectious, population and cases
  output = cbind(inf,
                 S = susceptibles$x, 
                 C = cases$cases.x, 
                 N = n$x)
  
  colnames(output)[c(3:8)]<- c("Group","Vaccinated","I","S","C","N")
  
  return(list(output,aggregate.data))
  
  
}

###Determine state based on model ####
SIR.state<- function(in.data,#vector of consecutive samples
                      model,#type of transmission model to use
                      inf.rule =1,
                      rec.rule =1,
                      only.last = T #if only last cell is positive remove
 ){
  out.data <- in.data;
  #determine potential infection and recovery events
  change <- c(0,in.data[2:length(in.data)]-in.data[1:(length(in.data)-1)])
    #then based on the model recode it to being susceptible (0), infectious (1) or recovered (2)

  if(model == "SI")
  {
    #first positive samples and loop such that infection rule is fullfilled
    found = F; index = 1;index.first <- length(change)+1
    while(index <= length(change) & found == F & max(change==1,na.rm = T) == 1 )
    {
    index.first <- c(1:length(out.data))[change==1][index]
    if(!is.na(index.first)){
      if(min(in.data[index.first:min(index.first+inf.rule-1, length(in.data))])==1){found = T}
      else
        {
        #false positive removal
        out.data[index.first]<- 0;
        #add 
        index = index + 1;}
      }else{index = index + 1;}
    }
    #get the NA's
    index.nas <- c(1:length(out.data))[is.na(out.data)]
    #set all to one
    if(!is.na(index.first)){if(index.first <= length(out.data)){out.data[index.first:length(out.data)] <- 1}}
    out.data[index.nas] <- NA
  }
  
  if(model == "SIR")
  {
    #first positive samples and loop such that infection rule is fullfilled
    found = F; index = 1;index.first <- length(change)+1
    while(index <= length(change) & found == F & max(change==1,na.rm = T) == 1)
    {
      index.first <- c(1:length(out.data))[change==1][index]
      if(!is.na(index.first)){
      if(min(in.data[index.first:min(index.first+inf.rule-1, length(in.data))])==1){
        found = T
        #remove changes before the first index 
        change[1:(index.first-1)] <- 0
        }
      else{
        #false positive removal
        out.data[index.first]<- 0;
        #add 
        index = index + 1;}
      }else{index = index + 1;}
    }
    
    #first negative samples and loop such that infection rule is fullfilled
    found = F; index = 1;index.last <- length(change)+1
    while(index <= length(change) & found == F)
    {
      index.last <- c(1:length(out.data))[change== -1 & !is.na(change)][index]
      if(!is.na(index.last))
      {
      if(max(in.data[index.last:min(index.last+rec.rule-1, length(in.data))])==0)
        {
        found = T
        }
      else{
        #false negative removal
        out.data[index.last]<- 1;
        #add 
        index = index + 1;}
      } else{index = index + 1}
    }
    #get the NA's
    index.nas <- c(1:length(out.data))[is.na(out.data)]
    if(is.na(index.first)|index.first> length(out.data))
    {
      index.first <- NA #no infection
      index.last<- NA #no recovery without infection
    }
    
    #set all to one
    if(!is.na(index.first)){
            out.data[index.first:length(out.data)] <- 1
    if(!is.na(index.last)){ out.data[(index.last):length(out.data)] <- 2}}
    out.data[index.nas] <- NA
    }
    
  
  #if(model == "SIS") {stop("not yet implemented")}
  
  if(only.last){
    if(sum(as.numeric(out.data[1,]),na.rm = T)==1){
      if(tail(out.data[!is.na(out.data)], n = 1) == 1)
        {
        out.data[!is.na(out.data)]<- 0}#set all to be 0
      }
      
  }
  
  return(out.data)
}





#standard error function
sem <- function(x){sd(x,na.rm =T)/sqrt(length(x[!is.na(x)]))}
lognorm.sem <- function(x){sum((log(x)- mean(log(x)))^2/length(x[!is.na(x)]),na.rm =T) / length(x[!is.na(x)])}
```

## Load and transform data

```{r loaddata}
#set directory with data
data.dir <- "C:/Surfdrive/Projecten/CEVA/"
#read the data of the vaccination group

#data original CT values
cevadata.original <- read.xlsx(paste0(data.dir, "DataNDVtrial_manual_Egil 20200317_Timea20200319.xlsx"),
                        sheet = "RAW (4)")

#data used for transmission 
cevadata <- read.xlsx(paste0(data.dir, "DataNDVtrial_manual_Egil 20200317_Timea20200319.xlsx"),
                        sheet = "Input_seedernotchanged")

#transform to usable data
data.one.SIR <- reform.SIR.data(data = cevadata,
                                sampleday.vars= paste0(c(1:14),".dpch"),
                                exclude.seeder = T )
data.one.SIR$Vaccinated <- factor(data.one.SIR$Vaccinated, levels = c("Yes","No"))
data.one.SIR$Vaccinated<-relevel(data.one.SIR$Vaccinated, ref = "No")

```

## Infectious period

#data selection

```{r Tinf, echo=TRUE}
#get data to calculate infectious period
out.indiv.one.SIR<- data.one.SIR[[2]];
#sum all positive intervals
out.indiv.one.SIR$infper <- apply(data.one.SIR[[2]][,c(5:18)]==1, 1, sum, na.rm = T);
#remove 0 as these are not infected thus should not be considered
out.indiv.one.SIR[out.indiv.one.SIR$infper == 0,]$infper<- NA #zero means never been infected
```

#plot

```{r include = F}
#make factor
out.indiv.one.SIR$Vaccinated<-factor(out.indiv.one.SIR$Vaccinated, levels = c("Yes","No"))

hist.one.chal <- ggplot(data = out.indiv.one.SIR) +
  geom_histogram(aes(x = infper, y = ..density.. , fill = Vaccinated), colour = "black",
                 position = "dodge",binwidth = 1, na.rm = T)+
  scale_fill_manual(values=c("Yes" = cbPalette[2],"No" = cbPalette[3]))+
  scale_x_continuous(breaks = c(0:10))+
  ggtitle("Distribution of infectious period")+
  ylab("Proportion of infected birds")+
  xlab("Infectious period (days)") +
  #theme(panel.background = NULL,  legend.position = "none")+
  theme_bw()+
  facet_grid( ~ Challenge,
             labeller = labeller(
       # Vaccinated = c("No" = "Unvaccinated", "Yes" = "Vaccinated"),
        Challenge = c("contact" = "Contact", "seeder" = "Seeder")
    )) 
#ggsave("InfectiousPeriodCEVA.tiff",hist.one.chal, dpi =720)

hist.one.chal
```

#estimate infectious period

The infectious period \(T_{inf}\): Parametric survival analyses assuming Normal distribution.

```{r}
#mixed effects model with group as random slope
fit.one.SIR.inf.mm<-lmer(infper ~ Vaccinated + Challenge + Challenge*Vaccinated+ (1|Group),data = out.indiv.one.SIR)

#fixed effects model 
fit.one.SIR.inf.fixed<-lm(infper ~ Vaccinated + Challenge + Challenge*Vaccinated,data = out.indiv.one.SIR)
#fixed effects model without interaction tem
fit.one.SIR.inf.fixed.no.int <- lm(infper ~ Vaccinated + Challenge ,data = out.indiv.one.SIR)

#Compare mixed-model to fixed effects model
anova(fit.one.SIR.inf.mm,fit.one.SIR.inf.fixed)
anova(fit.one.SIR.inf.fixed.no.int,fit.one.SIR.inf.fixed)
```
Conclusion a fixed effect model without interaction term can be used to estimate the infectious period.

```{r include = F}

#variable selection
drop1(fit.one.SIR.inf.fixed.no.int, test = "F", ~.+Vaccinated + Challenge )

```
Vaccination should be included in the model as well as challenge.

# Parameteric survival analyses 
```{r include =T}
#add censoring as variable to the data set
out.indiv.one.SIR$cens <- 1-as.numeric(out.indiv.one.SIR$`14.dpch` == 1)
out.indiv.one.SIR$cens[is.na(out.indiv.one.SIR$cens)] <- 1

#Survival analyses on contact birds
#perform exponential survival analysis
surv.fit.one.SIR.exponential <- survival::survreg(survival::Surv(infper,cens)~ Vaccinated, dist = "exponential",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)
#perform gaussian survival analysis 
surv.fit.one.SIR.gaussian <- survival::survreg(survival::Surv(infper,cens)~ Vaccinated , dist = "gaussian",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)
AIC(surv.fit.one.SIR.exponential)
AIC(surv.fit.one.SIR.gaussian)
```

Based on AIC select gaussian model.

```{r}
#check if dropping vaccination changes the model
drop1(surv.fit.one.SIR.gaussian,test = "Chisq")

#print confidence intervals
summary(surv.fit.one.SIR.gaussian)
Tinfci<-confint(surv.fit.one.SIR.gaussian)
summary(surv.fit.one.SIR.gaussian)$coefficients[1]
Tinfci[1,]
sum(summary(surv.fit.one.SIR.gaussian)$coefficients)
Tinfci[2,]+summary(surv.fit.one.SIR.gaussian)$coefficients[1]
#



```

drop1(surv.fit.one.SIR.gaussian,test = "Chisq")
summary(surv.fit.one.SIR.gaussian.sc)

#Calculate beta and R0

```{r}
#pre select data without rows for S =0 or I = 0 or S = NA or C = NA
out.nona.one.SIR <- data.one.SIR[[1]][(!is.na(data.one.SIR[[1]]$S)&!is.na(data.one.SIR[[1]]$C) &data.one.SIR[[1]]$I>0 &data.one.SIR[[1]]$S>0),]
out.nona.one.SIR$Vaccinated <- factor(out.nona.one.SIR$Vaccinated, levels = c("No","Yes"))
out.nona.one.SIR$Vaccinated <- relevel(out.nona.one.SIR$Vaccinated, ref = "No")
#estimate beta 
fit.SIR <- glm(cbind(C, S-C) ~ Vaccinated ,
                offset = log(I/N),
                family = binomial(link = "cloglog"), data = out.nona.one.SIR, 
                na.action = na.omit)
# #profile CI
# fit.SIR.confint <- confint(fit.SIR)
# summary(fit.SIR)
# beta.ci <- confint(fit.SIR)
# exp(fit.SIR$coefficients[1])
# exp(beta.ci[1,])
# exp(sum(fit.SIR$coefficients))
# exp(beta.ci[2,])

#calculate R0
alpha = 0.05 #conf.level
#beta
mean.logbeta <- summary(fit.SIR)$coefficients[,1]
confint.logbeta<-  confint(fit.SIR)

#infectious period
mean.inf.per <- summary(surv.fit.one.SIR.gaussian)$coefficients[1:2]
confint.inf.per <- confint(surv.fit.one.SIR.gaussian)[1:2,]
confint.inf.per[2,]<- confint.inf.per[1, ] +confint.inf.per[2, ]

#log infectious period
mean.log.inf.per <- log(cumsum(mean.inf.per))
confint.log.inf.per <- log(confint.inf.per)

#RNot rigth because of deifferent ordering.
a lskf adsf;j
mean.beta.inf.R <- data.frame(Vaccinated = c("No","Yes"),
                      beta = round(exp(cumsum(mean.logbeta)),2),
                      llbeta = round(exp(confint.logbeta[,1]+c(0,mean.logbeta[1])),2),
                      ulbeta = round(exp(confint.logbeta[,2]+c(0,mean.logbeta[1])),2),
                      infper = round(cumsum(mean.inf.per),2),
                      llinfper = round(confint.inf.per[,1],2),
                      ulinfper = round(confint.inf.per[,2],2),
                      logR0 = round(cumsum(mean.logbeta)+mean.log.inf.per,2),
                      lllogR0 = round(cumsum(mean.logbeta)  + confint.logbeta[,1] + log(confint.inf.per[,1]),2),
                      ullogR0 = round(cumsum(mean.logbeta)  + confint.logbeta[,2] + log(confint.inf.per[,2]),2)
)


mean.beta.inf.R$R <- round(exp(mean.beta.inf.R$logR0),2)
mean.beta.inf.R$llR <- round(exp(mean.beta.inf.R$lllogR0),2)
mean.beta.inf.R$ulR <- round(exp(mean.beta.inf.R$ullogR0),2)
rownames(mean.beta.inf.R)<- NULL
mean.beta.inf.R[, c(1,2,3,4)]
mean.beta.inf.R[, c(1,11,12,13)]

```

