---
title: "NDV_supplement with `A vector vaccine reduces transmission of Newcastle disease virus in commercial broiler chickens with maternally derived antibodies' "
author: "T. Tatar-Kis, E.A.J. Fischer, C. Cazaban, E. Kovacs, Z. G. Homonnay, V. Palya and J.A.(Arjan) Stegeman"
date: "24-8-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Used libraries and functions

#libraries

```{r}
options("getSymbols.warning4.0"=FALSE)
#add libraries
library(openxlsx)
library(ggplot2)
library(lme4)
library(survival)
```

# User defined functions

```{r}

#colour-blind colour palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

#function to reform data for glm
source("NDV_datamodel.R")

```

## Load and transform data

```{r loaddata}
#set directory with data equal to the working directory
data.dir <- getwd()
#read the data of the vaccination group

#data original CT values
ndvdata <- read.xlsx("Appendix_individualsCtvaluesandevaluation.xlsx", 
                 startRow =2,
                 colNames = T)

#cevadata.original <- read.xlsx(paste0(data.dir, "Appendix-individuals Ct values and evaluation.xlsx"),
 #                       sheet = "RAW (4)")
ndvdata<-fill.type.na.rename(ndvdata)
#data used for transmission 
#cevadata <- read.xlsx(paste0(data.dir, "DataNDVtrial_manual_Egil 20200317_Timea20200319.xlsx"),
 #                       sheet = "Input_seedernotchanged")

#transform to usable data
data.one.SIR <- reform.SIR.data(data = ndvdata,
                                sampleday.vars= paste0(c(1:14),".dpch"),
                                exclude.seeder = T )
data.one.SIR$Vaccinated <- factor(data.one.SIR$Vaccinated, levels = c("Yes","No"))
data.one.SIR$Vaccinated<-relevel(data.one.SIR$Vaccinated, ref = "No")

```

## Survival analyses
Determine if there is a difference in survival between vaccinated and unvaccinated birds.

```{r include = T}
#determine length of life
find.survival.time <- function(m){j = length(m);#look for last entry with value starting at end of vector
while(j > 0){
  if(!is.na(m[j]))return(j)
  j = j - 1
          }
}
out.indiv.one.SIR$survivaltime <-sapply(X = c(1:length(out.indiv.one.SIR$Vaccinated)), FUN = function(x){find.survival.time(out.indiv.one.SIR[x,6:19])} )
#add censoring as variable to the data set
out.indiv.one.SIR$event <- as.numeric(out.indiv.one.SIR$`14.dpch` == 1) #0 is alive
out.indiv.one.SIR$event[is.na(out.indiv.one.SIR$event)] <- 1 #NA is death
out.indiv.one.SIR[out.indiv.one.SIR$event==0,]

#survival analyses
surv.cont <-coxph(formula = Surv(survivaltime,event) ~ ci + Vaccinated, data = out.indiv.one.SIR)
surv.cont
drop1(surv.cont)
"Hazard ratio cannot be calculated because all vaccinated birds survive"

surv.cont <- coxph(formula = Surv(survivaltime,event) ~ Vaccinated , data = out.indiv.one.SIR[out.indiv.one.SIR$ci == "contact", ])
surv.cont
drop1(surv.cont)
"Hazard ratio cannot be calculated well because all vaccinated birds survive"

```

## Excretion
For those that were defined positive

```{r include = T}
ct.data <- ndvdata
for(day in c(6:19)){
  
  ct.data[(!data.one.SIR[[2]][,day]==1)|is.na(data.one.SIR[[2]][,day]),day]<- NA
  
}

suppressWarnings(
 ct.data$AUC <- sapply(c(1:length(ct.data$Group)), function(j){sum(36-as.numeric(ct.data[j,6:19]), na.rm = T)}) )#do not bother about NA's

#selection only the ct values >0
ct.data<- ct.data[ct.data$AUC>0,]
 
ctauc <- aggregate(formula = AUC ~ Vaccinated+ci+Sample,
          data = ct.data[ct.data$AUC>0,],
          FUN = "mean", rm.na = T)
ctauc$sd<-aggregate(formula = AUC ~ Vaccinated+ci+Sample,
          data = ct.data[ct.data$AUC>0,],
          FUN = "sd")$AUC
ctauc$n<-aggregate(formula = AUC ~ Vaccinated+ci+Sample,
          data = ct.data[ct.data$AUC>0,],
          FUN = "length")$AUC
#combine all categories
ct.data$category <- paste0(ct.data$Vaccinated,".",ct.data$ci)

#overall test for AUC against either one of the categories (categories is combination type bird and vaccination status)
kruskal.test(formula = AUC ~ category,data = ct.data[ct.data$Sample == "ON swab",])
kruskal.test(formula = AUC ~ category,data = ct.data[ct.data$Sample == "CLO swab",])
```

Compare challenge type (seeder or contact) stratisfied by type of swab.

```{r}
#cloacal swabs contact versus seeders
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "CLO swab",], paired = F)

#oro-nasal swabs contact versus seeders
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "ON swab",], paired = F)
```
Compare challenge type (seeder or contact) stratisfied by type of swab in unvaccinated birds.

```{r}
#test  seed=contact CLO in unvaccinated
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$Vaccinated =="No",], paired = F)
 
#test  seed=contact ON in unvaccinated
wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "ON swab"&ct.data$Vaccinated =="No",], paired = F)
```

Compare challenge type (seeder or contact) stratisfied by type of swab in vaccinated birds.

```{r} 
#test seed - contact per treatment CLO in vaccinated
 wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$Vaccinated =="Yes",], paired = F)
 ct.data[ct.data$Sample == "CLO swab"&ct.data$Vaccinated =="No"&ct.data$ci=="seeder",]$AUC
 ct.data[ct.data$Sample == "CLO swab"&ct.data$Vaccinated =="No"&ct.data$ci=="contact",]$AUC
 #test seed - contact per treatment ON in vaccinated
 wilcox.test(formula = AUC ~ Challenge,data = ct.data[ct.data$Sample == "ON swab"&ct.data$Vaccinated =="Yes",], paired = F)
  ct.data[ct.data$Sample == "ON swab"&ct.data$Vaccinated =="No"&ct.data$ci=="seeder",]$AUC
 ct.data[ct.data$Sample == "ON swab"&ct.data$Vaccinated =="No"&ct.data$ci=="contact",]$AUC

```

Compare vaccinated to unvaccinated  stratisfied by type of swab.

```{r}


#test vaccinated vs unvaccinated overall
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "CLO swab",])
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "ON swab",])


```


Compare vaccinated to unvaccinated  stratisfied by type of swab for contact birds .

```{r}

#test vaccinated versus unvaccinated for contact

wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$ci =="contact",])
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "ON swab"&ct.data$ci =="contact",]) 

#test vaccinated versus unvaccinated for seeder
wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "CLO swab"&ct.data$ci =="seeder",])

wilcox.test(formula = AUC ~ Vaccinated,data = ct.data[ct.data$Sample == "ON swab"&ct.data$ci =="seeder",])


```




## Infectious period

# Data selection

```{r Tinf, echo=TRUE}
#get data to calculate infectious period
out.indiv.one.SIR<- data.one.SIR[[2]];
#sum all positive intervals
out.indiv.one.SIR$infper <- apply(data.one.SIR[[2]][,c(5:18)]==1, 1, sum, na.rm = T);
#remove 0 as these are not infected thus should not be considered
out.indiv.one.SIR[out.indiv.one.SIR$infper == 0,]$infper<- NA #zero means never been infected
```

# Plot

```{r include = T}
#make factor
out.indiv.one.SIR$Vaccinated<-factor(out.indiv.one.SIR$Vaccinated, levels = c("Yes","No"))

hist.one.chal <- ggplot(data = out.indiv.one.SIR) +
  geom_histogram(aes(x = infper, y = ..density.. , fill = Vaccinated), colour = "black",
                 position = "dodge",binwidth = 1, na.rm = T)+
  scale_fill_manual(values=c("Yes" = cbPalette[2],"No" = cbPalette[3]))+
  scale_x_continuous(breaks = c(0:10))+
 # ggtitle("Distribution of infectious period")+
  ylab("Proportion of excreting chickens")+
  xlab("Infectious period (days)") +
  #theme(panel.background = NULL,  legend.position = "none")+
  theme_bw()+
  facet_grid( ~ Challenge,
             labeller = labeller(
       # Vaccinated = c("No" = "Unvaccinated", "Yes" = "Vaccinated"),
        Challenge = c("contact" = "Contact", "seeder" = "Seeder")
    )) 
ggsave("InfectiousPeriod.tiff",hist.one.chal, dpi =720)

hist.one.chal
```

# Estimate infectious period
The infectious period \(T_{inf}\): Parametric survival analyses assuming Normal distribution.

```{r include =T}
#add censoring as variable to the data set
out.indiv.one.SIR$cens <- 1-as.numeric(out.indiv.one.SIR$`14.dpch` == 1)
out.indiv.one.SIR$cens[is.na(out.indiv.one.SIR$cens)] <- 1
out.indiv.one.SIR[out.indiv.one.SIR$cens==0,]
#Survival analyses on contact birds
#perform exponential survival analysis on infectious period
surv.fit.one.SIR.exponential <- survival::survreg(survival::Surv(infper,cens)~ relevel(Vaccinated, ref= "No"), dist = "exponential",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)

#perform gaussian survival analysis on infectious period
surv.fit.one.SIR.gaussian <- survival::survreg(survival::Surv(infper,cens)~ relevel(Vaccinated, ref= "No") , dist = "gaussian",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)
AIC(surv.fit.one.SIR.exponential)
AIC(surv.fit.one.SIR.gaussian)
```

Based on AIC select gaussian model.

```{r}
#check if dropping vaccination changes the model
drop1(surv.fit.one.SIR.gaussian,test = "Chisq")

#print confidence intervals
summary(surv.fit.one.SIR.gaussian)
Tinfci<-confint(surv.fit.one.SIR.gaussian)

```

drop1(surv.fit.one.SIR.gaussian,test = "Chisq")
summary(surv.fit.one.SIR.gaussian.sc)

# Calculate beta and R0
```{r}
#pre select data without rows for S =0 or I = 0 or S = NA or C = NA
out.nona.one.SIR <- data.one.SIR[[1]][(!is.na(data.one.SIR[[1]]$S)&!is.na(data.one.SIR[[1]]$C) &data.one.SIR[[1]]$I>0 &data.one.SIR[[1]]$S>0),]
out.nona.one.SIR$Vaccinated <- factor(out.nona.one.SIR$Vaccinated, levels = c("No","Yes"))
out.nona.one.SIR$Vaccinated <- relevel(out.nona.one.SIR$Vaccinated, ref = "No")

#add the presence of susceptibles in the seeder groups of the vaccinated
out.nona.one.SIR[out.nona.one.SIR$Group == "1A",]$S <- out.nona.one.SIR[out.nona.one.SIR$Group == "1A",]$S + 1
out.nona.one.SIR[out.nona.one.SIR$Group == "1B",]$S <- out.nona.one.SIR[out.nona.one.SIR$Group == "1B",]$S + 3

#estimate beta 
fit.SIR <- glm(cbind(C, S-C) ~ relevel(Vaccinated,ref = "No") ,
                offset = log(I/N),
                family = binomial(link = "cloglog"), data = out.nona.one.SIR, 
                na.action = na.omit)
#calculate R0
alpha = 0.05 #conf.level
#beta
mean.logbeta <- cumsum(summary(fit.SIR)$coefficients[,1])
var.logbeta <- summary(fit.SIR)$coefficients[,2]^2
confint.logbeta<-  confint(fit.SIR)+c(0,mean.logbeta[1])

#infectious period
mean.inf.per <- cumsum(summary(surv.fit.one.SIR.gaussian)$coefficients[1:2])
confint.inf.per <- confint(surv.fit.one.SIR.gaussian)[1:2,]+c(0, mean.inf.per[1])
#get the variance from the profile likelihood
var.inf.per <- ((confint.inf.per-mean.inf.per)[,2]/qnorm(0.05/2))^2
#calculate log of the mean and the variance of the log of the mean. 
mean.log.inf.per <- log(mean.inf.per)
var.log.inf.per <- log(1+var.inf.per/(mean.inf.per)^2)

#calculate R
mean.logR <- mean.logbeta + mean.log.inf.per
var.logR <- var.logbeta + var.log.inf.per
cll.logR <- mean.logR - qnorm(1-0.05/2) * sqrt(var.logR)
cul.logR <- mean.logR + qnorm(1-0.05/2) * sqrt(var.logR)

#put an a table
data.table <- data.frame(Tinf = round(mean.inf.per,2),
          Tinfll = round(confint.inf.per[,1],2),
          Tinful = round(confint.inf.per[,2],2),
          beta = round(exp(mean.logbeta),2),
          betall = round(exp(confint.logbeta)[,1],2),
          betaul =round(exp(confint.logbeta)[,2],2),
          R0 = round(exp(mean.logR),2),
          R0ll = round(exp (cll.logR),2),
          R0ul = round(exp(cul.logR),2))
row.names(data.table)<- c("Unvaccinated","Vaccinated")
out.table <- knitr::kable(x = data.table,
             digits =2,
             format = "html",
             #row.names = c("Unvaccinated","Vaccinated")),
             col.names =c("Estimate","2.5%","97.5%",
                          "Estimate","2.5%","97.5%",
                          "Estimate","2.5%","97.5%"))
kableExtra::add_header_above(out.table,c("Treatment" =1,
                 "Tinf" = 3,
                 "beta" = 3,
                 "R0" = 3))


```

Probability of a major outbreak is 0 for $R_0<= 1$ and $1-1/R_0$ otherwise. This is an approximation, because the infectious period is normally distributed. 

```{r}
#approximate probability major outbreak
(1-1/data.table$R0 ) #negative means 0
(1-1/data.table$R0ll ) #negative means 0
(1-1/data.table$R0ul ) #negative means 0

```

The final size can still be calculated as the absolute number of infections by one initial infected $\dfrac{1}{1-R_0}$ or the fraction of the population $z$ eventually infected $-\ln(1-z) = z R_0$ :

```{r}
#approximate final size of the outbreak
final.size<- function(R0){if(R0<1){return(1/(1-R0))}else{ 
 return(uniroot(function(z){-R0*z-log(1-z)},lower = .1, upper =1)$root)
  }
}
sapply(data.table$R0,final.size) #first is fraction, second is number
sapply(data.table$R0ll,final.size) #first is fraction, second is number
sapply(data.table$R0ul,final.size) #both fractions


```

# Sensitivity to assumptions on non-excreting vaccinated seeder birds
Three birds in the vaccinated group were challenged, but never started to excreet. In the estimates above we have assumed that these birds were excluded completely from the analyses. We will account for these birds in two ways in additional analyses: 
- The birds are neither susceptible nor infectious, but do add to the total number of birds (i.e. assuming these birds are immune)
- Exclude these birds completely.

```{r include = F}
#subtract 1 in group 1A and 3 in group 1B from susceptible to assume these birds are immune 
out.nona.one.SIR.sensR <- out.nona.one.SIR
out.nona.one.SIR.sensR[out.nona.one.SIR.sensR$Group == "1A",]$S <- out.nona.one.SIR.sensR[out.nona.one.SIR.sensR$Group == "1A",]$S - 1
out.nona.one.SIR.sensR[out.nona.one.SIR.sensR$Group == "1B",]$S <- out.nona.one.SIR.sensR[out.nona.one.SIR.sensR$Group == "1B",]$S - 3


fit.SIR.sensR <- glm(cbind(C, S-C) ~ relevel(Vaccinated,ref = "No") ,
                offset = log(I/N),
                family = binomial(link = "cloglog"), data = out.nona.one.SIR.sensR,
                na.action = na.omit)
#beta and R0
mean.logbeta.sensR <- cumsum(summary(fit.SIR.sensR)$coefficients[,1])
var.logbeta.sensR <- summary(fit.SIR.sensR)$coefficients[,2]^2
confint.logbeta.sensR<-  confint(fit.SIR.sensR)+c(0,mean.logbeta.sensR[1])

#calculate R
mean.logR.sensR <- mean.logbeta.sensR + mean.log.inf.per
var.logR.sensR <- var.logbeta.sensR + var.log.inf.per
cll.logR.sensR <- mean.logR.sensR - qnorm(1-0.05/2) * sqrt(var.logR.sensR)
cul.logR.sensR <- mean.logR.sensR + qnorm(1-0.05/2) * sqrt(var.logR.sensR)


```
```{r}
#summary of each fit
summary(fit.SIR)
summary(fit.SIR.sensR)


#confidence intervals of beta and R of each fit
exp(mean.logbeta.sensR)
exp(confint.logbeta.sensR)

exp(mean.logR.sensR)
exp(c(cll.logR.sensR , cul.logR.sensR))


```
