---
title: "NDVtrial_presentation_2"
author: "Egil Fischer"
date: "17-3-2020"
output:
  ioslides_presentation: default
  beamer_presentation: 
   theme: Berkeley
   slide_level: 2
  powerpoint_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
#add libraries
library(openxlsx)
library(ggplot2)
library(lme4)
#source functions
source("TransmissionExperimentDatamodel.R")
#standard error function
sem <- function(x){sd(x,na.rm =T)/sqrt(length(x[!is.na(x)]))}
lognorm.sem <- function(x){sum((log(x)- mean(log(x)))^2/length(x[!is.na(x)]),na.rm =T) / length(x[!is.na(x)])}

```

# Introduction
## Trial set-up
NDV trial:

- Transmission in vaccinated and unvaccinated groups

- Four separate groups (i.e. 2 replicates per treatment)

- Challenge in 10 out of 20 birds

- ON- and CLO-swab at each day post challenge (DPI) for 14 days

- ELISA at the end of the trial




## Transmission parameteres
\(R = \beta \cdot T_{inf}\)

- Infectious period \(T_{inf}\) = number of days a bird is infectious

- Transmission coefficient \(\beta\) = number of new infections by one infectious bird per day 

- Reproduction ratio \(R\) = number of new infectious birds per infectious bird during its entire infectious period


If \(R < 1\) an infection will go extinct.

## Research questions
- What are the values of the transmission parameters?
- Do the transmission parameters differ between vaccinated and unvaccinated groups?
- Is the \(R\) in vaccinated groups below 1?


<!-- ## Data -->
<!-- Four different assumptions were used to estimate the transmission parameter. -->
<!-- An animal is positive if: -->
<!-- \begin{enumerate} -->
<!-- \item one or both samples are positive -->
<!-- \item both samples are positive -->
<!-- \item the ON sample is positive -->
<!-- \item the CLO sample is positive -->
<!-- \end{enumerate} -->
```{r include = F }
data.dir <- "C:/Surfdrive/Projecten/CEVA/"
#read the data of the vaccination group
cevadata1A <- read.xlsx(paste0(data.dir,
"P062VectormuneNDtransmissionstudysheddingdata_setA_readible.xlsx"),
                      sheet = "Group1A")
cevadata2A <- read.xlsx(paste0(data.dir, "P062VectormuneNDtransmissionstudysheddingdata_setA_readible.xlsx"),
                        sheet = "Group2A")
cevadata1B <- read.xlsx(paste0(data.dir,
"P062VectormuneNDtransmissionstudysheddingdata_setB_readible.xlsx"),
                      sheet = "Group1B")
cevadata2B <- read.xlsx(paste0(data.dir, "P062VectormuneNDtransmissionstudysheddingdata_setB_readible.xlsx"),
                        sheet = "Group2B")

cevadata <- rbind(cevadata1A,cevadata1B,cevadata2A,cevadata2B)

#rename to standard
cevadata$ci <- sapply(cevadata$Challenge, FUN = function(x){ifelse(grepl("contact", x),"contact","seeder")})

```
# Material and methods
## Visualization of descriptive data
- Survival plot

- Epidemic curves (number of infectious animals divided by the total) were produced.

- Histogram of infectious period

## Statistical analyses
The infectious period \(T_{inf}\): Parametric survival analyses assuming Normal distribution

The transmission parameter \(\beta\): Generalized linear model with a complementary loglog-link function:

`glm(cbind(C, S-C) ~ 1,offset = log(I/N), family = binomial(link = "cloglog"), data = data.set, na.action = na.omit)`

The basic reproduction number \(\textit{R}_0 = \beta \cdot T_{inf}\)
The confidence interval is calculated by $\log(R) = \log(\beta)+\log(T_{inf})\pm Z_{0.05} \cdot (\log(SE(\beta)) +\log(SE(T_{inf}))$

Assumes independence between the transmission parameter and the infectious period.

## Transmission model and further assumptions
Transmission model

- SIR: Susceptible \(\rightarrow\) Infectious\(\rightarrow\) Recovered and immune

<!-- - SI: Susceptible \(\rightarrow\) Infectious until end of experiment -->

The determination of a single sample to be positive depends on two assumptions

- Detection threshold (default set to Ct < 36)

- Sample is positive if at least one consecutive sample is positive:  Pos , Neg , Pos, Pos = first positive is assumed to be negative, 3rd and 4th positive (same for negative but the opposite)

Or based on decision of Timea and Egil

# Results

```{r, include = FALSE}
data.one.SIR <- reform.data(cevadata,
                                sampleday.vars = paste0(c(1:14),".dpch"),
                                model = "SIR",
                                positive.if = "max",
                                inf.rule = 2,
                                rec.rule = 2,
                                cut.off = 36)

epicurveplot <- ggplot(data = cbind(data.one.SIR[[1]], pos = "At least one sample positive" )) +
  geom_path(aes(x = ndpch, y = I/N, colour = Vaccinated, group = Group))+
  geom_point(aes(x = ndpch, y = I/N, shape =  Group))+
  xlab("Day post inoculation")+ylab("Prevalence (I/N)")

survivalplot <- ggplot(data = data.one.SIR[[1]]) +
  geom_path(aes(x = ndpch, y = N, colour = Vaccinated, group = Group))+
  geom_point(aes(x = ndpch, y = N, shape =  Group))+
  xlab("Day post inoculation")+ylab("Number of chickens (N)")


data.ON.positive <- reform.data(cevadata[cevadata$Sample == "ON swab",],
                                sampleday.vars = paste0(c(1:14),".dpch"),
                                positive.if = "max",
                                cut.off = 36,
                                model = "SIR",
                                inf.rule = 2,
                                rec.rule =2)
data.CLO.positive <- reform.data(cevadata[cevadata$Sample == "CLO swab",],
                                sampleday.vars = paste0(c(1:14),".dpch"),
                                positive.if = "max",
                                cut.off = 36,
                                model = "SIR",
                                inf.rule = 2,
                                rec.rule =2)

epicurveplotONCLO <- ggplot(data = rbind(cbind(data.ON.positive[[1]], pos = "ON swab"),cbind(data.CLO.positive[[1]], pos = "CLO swab" ))) +
  geom_path(aes(x = ndpch, y = I/N, colour = Vaccinated, group = Group))+
  geom_point(aes(x = ndpch, y = I/N, shape =  Group))+
  facet_grid(.~pos)+xlab("Day post inoculation")+ylab("Prevalence (I/N)")

```

```{r include = F}
data.one.SIR.selection <- reform.data(cevadata,
                                sampleday.vars = paste0(c(1:14),".dpch"),
                                model = "SIR",
                                positive.if = "max",
                                inf.rule = 2,
                                rec.rule = 2,
                                cut.off = 36)
```

## Survival curve
```{r, fig.cap = 'Infection curves'}
survivalplot
```
## Epidemic curve SIR and one or both positive
```{r, fig.cap = 'Infection curves'}
epicurveplot
```
## Epidemic curve SIR and one or both positive
```{r, fig.cap = 'Infection curves'}
epicurveplotONCLO
```
## Further analysis based on epidemic curves

Further analyses will be presented for:

- One or both positive (at least one sample being positive)




<!--  Infectious period -->
```{r include = F}
#creat a dataframe to collect results
all.results <- NULL

```

```{r, include = F}
#get data to calculate infectious period
out.indiv.one.SIR<- data.one.SIR[[2]];
out.indiv.one.SIR$infper <- apply(data.one.SIR[[2]][,c(5:18)]==1, 1, sum, na.rm = T);
out.indiv.one.SIR[out.indiv.one.SIR$infper == 0,]$infper<- NA #zero means never been infected

out.indiv.one.SIR[out.indiv.one.SIR$Vaccinated =="Yes",c("Vaccinated","Challenge","infper")]
```

## \(T_{inf}\) at least one (CLO or ON) positive sample
```{r include = F}
#visualize
hist.one <- ggplot(data = out.indiv.one.SIR) +
  geom_histogram(aes(x = infper, y = ..density.., fill = Vaccinated),
                 position = "dodge",binwidth = 1, na.rm = T)+
  ggtitle("Distribution of infectious period with at least one sample pos")+
  xlab("Infectious period")
```

```{r}
hist.one
```

## \(T_{inf}\) at least one (CLO or ON) positive sample
```{r include = F}
hist.one.chal <- ggplot(data = out.indiv.one.SIR) +
  geom_histogram(aes(x = infper, y = ..density.., fill = Vaccinated),
                 position = "dodge",binwidth = 1, na.rm = T)+
  ggtitle("Distribution of infectious period with at least one sample pos")+
  xlab("Infectious period") +facet_grid(.~Challenge)
```

```{r}
hist.one.chal
```

<!-- ## Statistical analyses ~ fixed vs random effects model -->
```{r include = F}

fit.one.SIR.inf.mm<-lmer(infper ~ Vaccinated + Challenge + Challenge*Vaccinated+ (1|Group),data = out.indiv.one.SIR)
out.indiv.one.SIR$OneGroup = "G"
fit.one.SIR.inf.fixed<-lm(infper ~ Vaccinated + Challenge + Challenge*Vaccinated,data = out.indiv.one.SIR)
fit.one.SIR.inf.fixed.no.int <- lm(infper ~ Vaccinated + Challenge ,data = out.indiv.one.SIR)
#Compare mixed-model to fixed effects model
anova(fit.one.SIR.inf.mm,fit.one.SIR.inf.fixed)
anova(fit.one.SIR.inf.fixed.no.int,fit.one.SIR.inf.fixed)

```

<!-- The random effect "Group" can be ignored and the interaction term "Challenge x Vaccinated" also. -->
<!-- ## Statistical analyses ~ fixed vs random effects model -->
```{r include = F}

#Compare mixed-model to fixed effects model
drop1(fit.one.SIR.inf.fixed.no.int, test = "F", ~.+Vaccinated + Challenge )

```
<!-- Vaccination should be included in the model as well as challenge. -->
## Statistical analyses ~ using parameteric survival analyses

```{r include =F}
out.indiv.one.SIR$cens <- 1-as.numeric(out.indiv.one.SIR$`14.dpch` == 1)
out.indiv.one.SIR$cens[is.na(out.indiv.one.SIR$cens)] <- 1
surv.fit.one.SIR.exponential <- survival::survreg(survival::Surv(infper,cens)~ Vaccinated, dist = "exponential",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)
surv.fit.one.SIR.gaussian <- survival::survreg(survival::Surv(infper,cens)~ Vaccinated , dist = "gaussian",data = out.indiv.one.SIR[out.indiv.one.SIR$Challenge == "contact",], na.action = na.omit)

surv.fit.one.SIR.gaussian.sc <- survival::survreg(survival::Surv(infper,cens)~ Vaccinated+Challenge, dist = "gaussian",data = out.indiv.one.SIR, na.action = na.omit)

drop1(surv.fit.one.SIR.gaussian.sc)
anova(surv.fit.one.SIR.gaussian,surv.fit.one.SIR.exponential, test ="Chisq")
drop1(surv.fit.one.SIR.gaussian,test = "Chisq")
summary(surv.fit.one.SIR.gaussian.sc)

```
Survival analyses assuming a Gaussian distributed \(T_{Inf}\) and only using the contact birds. 
```{r}
summary(surv.fit.one.SIR.gaussian)
```


## Transmission parameter \(\beta\)
```{r include = F}
out.nona.one.SIR <- data.one.SIR[[1]][(!is.na(data.one.SIR[[1]]$S)&!is.na(data.one.SIR[[1]]$C) &data.one.SIR[[1]]$I>0 &data.one.SIR[[1]]$S>0),]

```

```{r include = F}
#GLM
fit.one.SIR.mm <- glmer(cbind(C, S-C) ~Vaccinated+
                  (1|Group),
                offset = log(I/N),
                family = binomial(link = "cloglog"), data = out.nona.one.SIR, na.action = na.omit)
fit.one.SIR.fixed <- glm(cbind(C, S-C) ~ Vaccinated ,
                offset = log(I/N),
                family = binomial(link = "cloglog"), data = out.nona.one.SIR, na.action = na.omit)

```
<!-- ## Compare model fits -->
```{r include = F}
#compare model with and without random effect. If there is no difference use the model without random effect
anova(fit.one.SIR.mm, fit.one.SIR.fixed)
```

<!-- ## Determine best model -->
```{r include = F}
drop1(fit.one.SIR.fixed)
summary(fit.one.SIR.fixed)
```


```{r, include = F}
fit.one.SIR.fixed.profile.confint <- confint(fit.one.SIR.fixed);
#beta's
beta.table<- data.frame(rbind(cbind(treatment = "Unvac",beta = round(exp(fit.one.SIR.fixed$coefficients[[1]]), digits =3), round(t(exp(fit.one.SIR.fixed.profile.confint[1,])),digits =3)),
cbind(treatment = "Vac", beta = round(exp(sum(fit.one.SIR.fixed$coefficients)),digits =3), round(t(exp(fit.one.SIR.fixed.profile.confint[2,]+fit.one.SIR.fixed$coefficients[[1]])),digits =3))));
colnames(beta.table)<- c("Treatment", "beta","2.5%","97.5")
```


```{r}
summary(fit.one.SIR.fixed)

```


## \(\beta\) estimates
```{r}
beta.table

```
Vaccination does not lower the infection parameter \(\beta\).

<!-- ## Basic reproduction number \(R_0\) -->
```{r, include = F}
#calculate R0
 alpha = 0.05 #conf.level
#beta
mean.logbeta <- summary(fit.one.SIR.fixed)$coefficients[,1]
confint.logbeta<- fit.one.SIR.fixed.profile.confint

#infectious period
mean.inf.per <- summary(surv.fit.one.SIR.gaussian)$coefficients[1:2]
confint.inf.per <- confint(surv.fit.one.SIR.gaussian)[1:2,]
confint.inf.per[2,]<- confint.inf.per[1, ] +confint.inf.per[2, ]

#log infectious period
mean.log.inf.per <- log(cumsum(mean.inf.per))
confint.log.inf.per <- log(confint.inf.per)

#R
mean.beta.inf.R <- data.frame(Vaccinated = c("No","Yes"),
                      beta = exp(cumsum(mean.logbeta)),
                      llbeta = exp(cumsum(confint.logbeta[,1])),
                      ulbeta = exp(cumsum(confint.logbeta[,2])),
                      infper = cumsum(mean.inf.per),
                      llinfper = confint.inf.per[,1],
                      ulinfper = confint.inf.per[,2],
                      logR0 = cumsum(mean.logbeta)+mean.log.inf.per,
                      lllogR0 = cumsum(mean.logbeta  + confint.logbeta[,1]) + log(confint.inf.per[,1]),
                      ullogR0 = cumsum(mean.logbeta  + confint.logbeta[,2]) + log(confint.inf.per[,2])
)


mean.beta.inf.R$R <- exp(mean.beta.inf.R$logR0)
mean.beta.inf.R$llR <- exp(mean.beta.inf.R$lllogR0)
mean.beta.inf.R$ulR <- exp(mean.beta.inf.R$ullogR0)
rownames(mean.beta.inf.R)<- NULL
```

## Summarize ~ infectious period
```{r}

mean.beta.inf.R[, c(1,5,6,7)]

```
Vaccine reduces length of infectious period (p = 0.004)

## Summarize ~ \(\beta\)
```{r}

mean.beta.inf.R[, c(1,2,3,4)]

```
No difference is transmission rate

## Summarize ~ \(R\)
```{r}

mean.beta.inf.R[, c(1,11,12,13)]

```
No difference in reproduction number

# Conclusion and discussion

## Conclusion
Based on the assumption of one of both samples required to be positive, we find that the transmission parameter \(\beta\) is not lower, but the mean infectious period \(T_{inf}\)in the contact birds is lower, which results in a reproduction number \(R\) which is  smaller (although not statistically significant), but is not lower than the threshold \(R \leq 1\).
<!-- ```{r include = F} -->
<!--  all.results <- rbind(all.results, cbind(Assumption = "OneOrBoth", mean.beta.inf.R)) -->

<!-- ``` -->

<!-- ## ON swabs -->

<!-- ## SI model -->

## Optional additional analyses
Preliminary results (models not checked)

- Different Ct as cut-off: e.g if Ct = 30, \(\beta\) lower values for vaccinated, \(T_{inf}\) cannot be estimated 


- One positive sample means that animal is positive and first negative is negative: \(\beta\) lower values for vaccinated, and \(T_{inf}\) lower for vaccinated

Following no preliminary results

- Only base results on ON swabs

- SI model

- Use ELISA to determine infection yes/no.





<!-- # Appendices -->
<!-- ## Model diagnostics -->

<!-- ## Model diagnostics -->

<!--## Compare raw data with reformed data-->
```{r include = F}

file.name <- paste0(format(Sys.Date(),"%y%m%d"),"compareOne.xlsx")
if(file.exists(file.name)) {file.remove(file.name)}
wb <- createWorkbook(creator = "Egil A. J. Fischer")
addWorksheet(wb,   sheetName = "RAW")
addWorksheet(wb,   sheetName = "Reform")
writeData(wb, sheet = "RAW",cevadata[order(cevadata$bird.id),], colNames = T)
writeData(wb, sheet = "Reform", out.indiv.one.SIR[order(out.indiv.one.SIR$bird.id),], colNames = T)
saveWorkbook(wb, file = file.name)

```


